Evolution of Processes in Delivering Reliable Software within the European Scientific Arena

### 2019-02-14 paper-jgc.tex
Take defects figure out from conclusions.
Add Pablo as first author
Use PID for sqa-baseline
Double-check online URLs and update access date

\author{M. David \and
        D. C. Duma \and
        P. Orviz Fernandez \and
        E. Ronchieri \and
        J. Gomes \and
        D. Salomoni
%
%First Author         \and
%        Second Author %etc.
}

@@

\institute{M. David \at
LIP, Lisbon, Portugal \\
%Tel.: +123-45-678910\\
%Fax: +123-45-678910\\
\email{david@lip.pt}           %  \\
%             \emph{Present address:} of F. Author  %  if needed
\and
D. C. Duma \at
INFN - CNAF, Bologna, Italy\\
\email{cristina.aiftimiei@cnaf.infn.it}
\and
P. Orviz Fernandez \at
CSIC, Santander, Spain \\
\email{orviz@ifca.unican.es}
\and
J. Gomes \at
LIP, Lisbon, Portugal \\
%Tel.: +123-45-678910\\
%Fax: +123-45-678910\\
\email{jorge@lip.pt}
\and
E. Ronchieri \at
INFN - CNAF, Bologna, Italy\\
\email{elisabetta.ronchieri@cnaf.infn.it}
\and
D. Salomoni \at
INFN - CNAF, Bologna, Italy\\
\email{davide.salomoni@cnaf.infn.it}
}

@@

\subsection{Software quality validation and delivery automation}
\label{sec:devops}
Tackling the granularity of the former SQA requirements -- most of which accounted in a per--change basis --
is hardly achieved without the aid of automation. The introduction of automation to validate the quality
requirements increases the overall reliability of the produced software: automated testing is more
time-efficient, leading to higher code coverages and increased defect detection, when compared with a manual
approach \cite{rafi2012}.

@@

% Comment
%%P8 C1 L26: What is the demonstration that reliability has been improved significantly with the addition of stricter methologies? Is there some quantitative evidence of uptime, or bug reports per user, or similar? Or some evidence of speed of defect detection & fixing, which you cite as greatly improved? I am very sympathetic to the idea that systematic development techniques can help improve software quality (without getting into circular definitions, where "quality" means that methodologies or standards have been applied!) but there will also be a point at which the methodology gets too heavy and compromises the "agility" of the development team: how can you identify which processes are helping and which are not?
Reliability of the software has been substantially improved by applying the SQA
criteria in the development stage. Consequently, each change in the source code
that is meant to be integrated in the production version, is validated according
to the SQA testing requirements. Thus, defects are detected and corrected early in the
software lifecycle, reducing the cost of fixing errors at later stages. Automation is
key in this process, allowing the implementation of a CI scenario that enforces
the SQA criteria compliance. The application of more advanced DevOps practices
resulted in the implementation of CD pipelines to extend automation to the
delivery of packages. Thereby, safe and frequent software releases are possible, reducing
the impact on production systems. In this regard, a pre--production validation,
through the preview testbeds or resource center's early adoption -- handled by the
Staged--rollout process -- alleviates the likelihood of discovering uncovered
issues at production. To put it in numbers, as Fig.~\ref{fig:indigobugs} showcases, the
total number of defects detected in the pre--production stages surpassed by a factor of 30
the ones reported by external users, over the lifetime of the {\sl INDIGO-DataCloud}
project.
\begin{figure}[h]
\centering
\includegraphics[width=0.8\textwidth, height=50mm]{images/indigoBugs.png}
\caption{Number of defects detected in development and production stages over the lifetime of the
{\sl INDIGO-DataCloud} project. The defects started to arise in the second year of the project,
in parallel with the release of {\sl INDIGO--1} software, while the major activity was concentrated in the
months that preceded the second major release of {\sl INDIGO--2}, in April 2017. The practical implementation
of the SQA definition described throughout this paper was the key actor in uncovering defects while in pre--production
stages, thus not affecting the software released.}
\label{fig:indigobugs}
\end{figure}

@@

\bibitem{h2020}
\emph{Horizon 2020} -- European Commission.
[Online]. Available: \url{https://ec.europa.eu/programmes/horizon2020/}.
[Accessed 4 Jun. 2018].

@@

\bibitem{agile-manifesto}
K. Beck \emph{et al.},
``Manifesto for Agile Software Development,'' 2012.
[Online]. Available: \url{http://www.agilemanifesto.org/}.
[Accessed 4 Jun. 2018].

@@

\bibitem{cmm}
M. Paulk, B. Curtis, M. Chrissis and V. C. Weber,
``Capability maturity model for software,''
Softw. Eng. Inst.,
Technical Report CMU/SEI-93-TR-024, ESC-TR-93-177, Feb. 1993.
[Online]. Available: \url{https://resources.sei.cmu.edu/asset_files/TechnicalReport/1993_005_001_16211.pdf}.
[Accessed 4 Jun. 2018].

\bibitem{datagridguide}
Quality Assurance Group, ``DataGrid - European DataGrid Developers' Guide,''
2003.
[Online] Available: \url{https://edms.cern.ch/ui/file/358824/1.1/EDG-DevGuide-v1-2.pdf}
[Accessed 30 Aug. 2018].

\bibitem{datagridindicators}
DataGrid, ``DataGrid Internal Document - Quality and Performance Indicators for DataGrid,''
2003.
[Online] Available: \url{https://edms.cern.ch/ui/file/386039/2/QIv0-3.pdf}
[Accessed 30 Aug. 2018].

\bibitem{cordis:egee}
\emph{Enabling Grids for E--sciencE (EGEE)} project, European Community Research and
Development Information Service (CORDIS).
[Online] Available: \url{http://cordis.europa.eu/project/rcn/80149_en.html}
[Accessed 4 Jun. 2018].

\bibitem{cordis:egee2}
\emph{Enabling Grids for E--sciencE--II (EGEE--II)} project, European Community Research and
Development Information Service (CORDIS).
[Online] Available: \url{http://cordis.europa.eu/project/rcn/99189_en.html}
[Accessed 4 Jun. 2018].

\bibitem{cordis:egee3}
\emph{Enabling Grids for E--sciencE--III (EGEE--III)} project, European Community
Research and Development Information Service (CORDIS).
[Online] Available: \url{http://cordis.europa.eu/project/rcn/87264_en.html}
[Accessed 4 Jun. 2018].

@@

\bibitem{egee:acceptance-criteria}
\emph{Definition and Documentation of the Revised Software Life--Cycle Process},
Milestone MSA3.4.2, 2010, EGEE--III project.
[Online]. Available: \url{https://edms.cern.ch/ui/file/1062487/2/EGEE-III-MSA3.4.2-1062487-v1_4.pdf}.
[Accessed 4 Jun. 2018].

@@

\bibitem{iso-9126}
\emph{ISO/IEC 9126 Software Engineering - Product Quality},
International Organization for Standardization.
[Online]. Available: \url{https://www.iso.org/standard/22749.html}.
[Accessed 4 Jun. 2018].

@@

\bibitem{egi-qc}
\emph{EGI Quality Criteria}.
[Online]. Available: \url{https://egi-qc.github.io/}.
[Accessed 4 Jun. 2018].

\bibitem{cordis:egi-engage}
\emph{Engaging the EGI Community towards an Open Science Commons (EGI--ENGAGE)}
project, European Community Research and Development Information Service (CORDIS).
[Online] Available: \url{http://cordis.europa.eu/project/rcn/194937_en.html}
[Accessed 4 Jun. 2018].

@@

\bibitem{indigo-d31}
J. Gomes \emph{et al.},
``Initial Plan for WP3,''
\emph{INDIGO--DataCloud Deliverable 3.1}.
[Online]. Available: \url{https://www.indigo-datacloud.eu/documents/initial-plan-wp3-d31}.
[Accessed 4 Jun. 2018].

\bibitem{sqa-baseline}
Pablo Orviz \emph{et al.},
``A set of common software quality assurance baseline criteria for research projects,''
2017.
[Online]. Available: \url{http://digital.csic.es/bitstream/10261/160086/4/CommonSQA-v2.pdf}.
[Accessed 4 Jun. 2018].

@@

\bibitem{foot1}
indigo-dc, ``Sign in to GitBook - GitBook,''
2018.
[Online]. Available: \url{https://www.gitbook.com/@indigo-dc}.
[Accessed 30 Aug. 2018].

\bibitem{foot2}
puppetforge, ``Modules by INDIGO Datacloud - Puppet Forge,''
2018.
[Online]. Available: \url{https://forge.puppet.com/indigodc}.
[Accessed 30 Aug. 2018].

\bibitem{foot3}
GALAXY, ``Ansible Galaxy,''
2018.
[Online]. Available: \url{https://galaxy.ansible.com/indigo-dc/}
[Accessed 30 Aug. 2018].

\bibitem{foot4}
GitHub Developer, ``GitHub API v3 | GitHub Developer Guide,''
2018.
[Online]. Available: \url{https://developer.github.com/v3/}
[Accessed 30 Aug. 2018].

\bibitem{foot5}
Jenkins, ``Jenkins,''
2018.
[Online]. Available: \url{https://jenkins.io/}
[Accessed 30 Aug. 2018].

\bibitem{foot6}
GRIMOIRELAB, ``GrimoireLab - Software Development and Community Analytics Platform,''
2017.
[Online]. Available: \url{http://grimoirelab.github.io/}
[Accessed 30 Aug. 2018].

\bibitem{foot7}
Jenkins Indigo-dc, ``Jenkins - Indigo-DataCloud,''
2018.
[Online]. Available: \url{https://jenkins.indigo-datacloud.eu/}
[Accessed 30 Aug. 2018].

\bibitem{foot8}
Indigo-dc, ``indigodatacloud - Docker Hub,''
2018.
[Online]. Available: \url{https://hub.docker.com/u/indigodatacloud}
[Accessed 30 Aug. 2018].

@@

\bibitem{foot9}
GitHub, ``Learn Git and GitHub without any code!,''
2018.
[Online]. Available: \url{https://github.com/}
[Accessed 30 Aug. 2018].

>>>>

\author{P. Orviz Fernandez \and
        M. David \and
        D. C. Duma \and
        E. Ronchieri \and
        J. Gomes \and
        D. Salomoni
%
%First Author         \and
%        Second Author %etc.
}

@@

\institute{
P. Orviz Fernandez \at
CSIC, Santander, Spain \\
\email{orviz@ifca.unican.es}
\and
M. David \at
LIP, Lisbon, Portugal \\
%Tel.: +123-45-678910\\
%Fax: +123-45-678910\\
\email{david@lip.pt}           %  \\
%             \emph{Present address:} of F. Author  %  if needed
\and
D. C. Duma \at
INFN - CNAF, Bologna, Italy\\
\email{cristina.aiftimiei@cnaf.infn.it}
\and
J. Gomes \at
LIP, Lisbon, Portugal \\
%Tel.: +123-45-678910\\
%Fax: +123-45-678910\\
\email{jorge@lip.pt}
\and
E. Ronchieri \at
INFN - CNAF, Bologna, Italy\\
\email{elisabetta.ronchieri@cnaf.infn.it}
\and
D. Salomoni \at
INFN - CNAF, Bologna, Italy\\
\email{davide.salomoni@cnaf.infn.it}
}

@@

\subsection{Software quality validation and delivery automation}
\label{sec:devops}
Tackling the granularity of the former SQA requirements -- most of which accounted in a per--change basis --
is hardly achieved without the aid of automation. The introduction of automation to validate the quality
requirements increases the overall reliability of the produced software: automated testing is more
time-efficient, leading to higher code coverages and increased defect detection, when compared with a manual
approach \cite{rafi2012}. To put it in numbers, as Fig.~\ref{fig:indigobugs} showcases, the
total number of defects detected in the pre--production stages surpassed by a factor of 30
the ones reported by external users, over the lifetime of the {\sl INDIGO-DataCloud}
project.
\begin{figure}[h]
\centering
\includegraphics[width=0.8\textwidth, height=50mm]{images/indigoBugs.png}
\caption{Number of defects detected in development and production stages over the lifetime of the
{\sl INDIGO-DataCloud} project. The defects started to arise in the second year of the project,
in parallel with the release of {\sl INDIGO--1} software, while the major activity was concentrated in the
months that preceded the second major release of {\sl INDIGO--2}, in April 2017. The practical implementation
of the SQA definition described throughout this paper was the key actor in uncovering defects while in pre--production
stages, thus not affecting the software released.}
\label{fig:indigobugs}
\end{figure}

@@

% Comment
%%P8 C1 L26: What is the demonstration that reliability has been improved significantly with the addition of stricter methologies? Is there some quantitative evidence of uptime, or bug reports per user, or similar? Or some evidence of speed of defect detection & fixing, which you cite as greatly improved? I am very sympathetic to the idea that systematic development techniques can help improve software quality (without getting into circular definitions, where "quality" means that methodologies or standards have been applied!) but there will also be a point at which the methodology gets too heavy and compromises the "agility" of the development team: how can you identify which processes are helping and which are not?
Reliability of the software has been substantially improved by applying the SQA
criteria in the development stage. Consequently, each change in the source code
that is meant to be integrated in the production version, is validated according
to the SQA testing requirements. Thus, defects are detected and corrected early in the
software lifecycle, reducing the cost of fixing errors at later stages. Automation is
key in this process, allowing the implementation of a CI scenario that enforces
the SQA criteria compliance. The application of more advanced DevOps practices
resulted in the implementation of CD pipelines to extend automation to the
delivery of packages. Thereby, safe and frequent software releases are possible, reducing
the impact on production systems. In this regard, a pre--production validation,
through the preview testbeds or resource center's early adoption -- handled by the
Staged--rollout process -- alleviates the likelihood of discovering uncovered
issues at production.

@@

\bibitem{h2020}
\emph{Horizon 2020} -- European Commission.
[Online]. Available: \url{https://ec.europa.eu/programmes/horizon2020/}.
[Accessed 14 Feb. 2019].

@@

\bibitem{agile-manifesto}
K. Beck \emph{et al.},
``Manifesto for Agile Software Development,'' 2012.
[Online]. Available: \url{http://www.agilemanifesto.org/}.
[Accessed 14 Feb. 2019].

@@

\bibitem{cmm}
M. Paulk, B. Curtis, M. Chrissis and V. C. Weber,
``Capability maturity model for software,''
Softw. Eng. Inst.,
Technical Report CMU/SEI-93-TR-024, ESC-TR-93-177, Feb. 1993.
[Online]. Available: \url{https://resources.sei.cmu.edu/asset_files/TechnicalReport/1993_005_001_16211.pdf}.
[Accessed 14 Feb. 2019].

\bibitem{datagridguide}
Quality Assurance Group, ``DataGrid - European DataGrid Developers' Guide,''
2003.
[Online] Available: \url{https://edms.cern.ch/ui/file/358824/1.1/EDG-DevGuide-v1-2.pdf}
[Accessed 14 Feb. 2019].

\bibitem{datagridindicators}
DataGrid, ``DataGrid Internal Document - Quality and Performance Indicators for DataGrid,''
2003.
[Online] Available: \url{https://edms.cern.ch/ui/file/386039/2/QIv0-3.pdf}
[Accessed 14 Feb. 2019].

\bibitem{cordis:egee}
\emph{Enabling Grids for E--sciencE (EGEE)} project, European Community Research and
Development Information Service (CORDIS).
[Online] Available: \url{http://cordis.europa.eu/project/rcn/80149_en.html}
[Accessed 14 Feb. 2019].

\bibitem{cordis:egee2}
\emph{Enabling Grids for E--sciencE--II (EGEE--II)} project, European Community Research and
Development Information Service (CORDIS).
[Online] Available: \url{http://cordis.europa.eu/project/rcn/99189_en.html}
[Accessed 14 Feb. 2019].

\bibitem{cordis:egee3}
\emph{Enabling Grids for E--sciencE--III (EGEE--III)} project, European Community
Research and Development Information Service (CORDIS).
[Online] Available: \url{http://cordis.europa.eu/project/rcn/87264_en.html}
[Accessed 14 Feb. 2019].

@@

\bibitem{egee:acceptance-criteria}
\emph{Definition and Documentation of the Revised Software Life--Cycle Process},
Milestone MSA3.4.2, 2010, EGEE--III project.
[Online]. Available: \url{https://edms.cern.ch/ui/file/1062487/2/EGEE-III-MSA3.4.2-1062487-v1_4.pdf}.
[Accessed 14 Feb. 2019].

@@

\bibitem{iso-9126}
\emph{ISO/IEC 9126 Software Engineering - Product Quality},
International Organization for Standardization.
[Online]. Available: \url{https://www.iso.org/standard/22749.html}.
[Accessed 14 Feb. 2019].

@@

\bibitem{egi-qc}
\emph{EGI Quality Criteria}.
[Online]. Available: \url{https://egi-qc.github.io/}.
[Accessed 14 Feb. 2019].

\bibitem{cordis:egi-engage}
\emph{Engaging the EGI Community towards an Open Science Commons (EGI--ENGAGE)}
project, European Community Research and Development Information Service (CORDIS).
[Online] Available: \url{http://cordis.europa.eu/project/rcn/194937_en.html}
[Accessed 14 Feb. 2019].

@@

\bibitem{indigo-d31}
J. Gomes \emph{et al.},
``Initial Plan for WP3,''
\emph{INDIGO--DataCloud Deliverable 3.1}.
[Online]. Available: \url{https://www.indigo-datacloud.eu/documents/initial-plan-wp3-d31}.
[Accessed 14 Feb. 2019].

\bibitem{sqa-baseline}
Pablo Orviz \emph{et al.},
``A set of common software quality assurance baseline criteria for research projects,''
2017.
[Online]. Available: \url{http://hdl.handle.net/10261/160086}.
[Accessed 14 Feb. 2019].

@@

\bibitem{foot1}
indigo-dc, ``Sign in to GitBook - GitBook,''
2018.
[Online]. Available: \url{https://www.gitbook.com/@indigo-dc}.
[Accessed 14 Feb. 2019].

\bibitem{foot2}
puppetforge, ``Modules by INDIGO Datacloud - Puppet Forge,''
2018.
[Online]. Available: \url{https://forge.puppet.com/indigodc}.
[Accessed 14 Feb. 2019].

\bibitem{foot3}
GALAXY, ``Ansible Galaxy,''
2018.
[Online]. Available: \url{https://galaxy.ansible.com/indigo-dc/}
[Accessed 14 Feb. 2019].

\bibitem{foot4}
GitHub Developer, ``GitHub API v3 | GitHub Developer Guide,''
2018.
[Online]. Available: \url{https://developer.github.com/v3/}
[Accessed 14 Feb. 2019].

\bibitem{foot5}
Jenkins, ``Jenkins,''
2018.
[Online]. Available: \url{https://jenkins.io/}
[Accessed 14 Feb. 2019].

\bibitem{foot6}
GRIMOIRELAB, ``GrimoireLab - Software Development and Community Analytics Platform,''
2017.
[Online]. Available: \url{http://grimoirelab.github.io/}
[Accessed 14 Feb. 2019].

\bibitem{foot7}
Jenkins Indigo-dc, ``Jenkins - Indigo-DataCloud,''
2018.
[Online]. Available: \url{https://jenkins.indigo-datacloud.eu:8080/}
[Accessed 14 Feb. 2019].

\bibitem{foot8}
Indigo-dc, ``indigodatacloud - Docker Hub,''
2018.
[Online]. Available: \url{https://hub.docker.com/u/indigodatacloud}
[Accessed 14 Feb. 2019].

@@

\bibitem{foot9}
GitHub, ``Learn Git and GitHub without any code!,''
2018.
[Online]. Available: \url{https://github.com/}
[Accessed 14 Feb. 2019].
